---
title: "Programming in Base R"
format: html
editor: visual
---

## Task 1: Basic Vector Practice

First we create two vectors, one corresponding to the pre-treatment measurements and one corresponding to the post-treatment measurements.

```{r}
pre_treat <- c(120,151,125,126,115,132,132,129,134,139,127,122,127,135,133,128,147,138,140,132)
pre_treat

post_treat <- c(127,145,135,122,115,122,123,126,126,129,132,146,120,114,121,120,128,120,133,115)
post_treat
```

Next we assign names to the each of the vectors elements.

```{r}
#names the ith element in vector x Subject_i
names(pre_treat) <- paste("Subject", 1:20, sep = '_') 

#names the ith element in vector y Subject_i
names(post_treat) <- paste("Subject", 1:20, sep = '_') 
```

Next we calculate the change in blood pressure observed for each patient after treatment.

```{r}
change_in_bp <- post_treat - pre_treat
change_in_bp
```

Now we calculate the average decrease in blood pressure across all patients.

```{r}
avg_change <- mean(change_in_bp)
avg_change
```

The average change in blood pressure is -5.65, so on average the patients saw a decrease in blood pressure (indicating that the treatment was effective). Next, we want to determine which specific patients saw a decrease in blood pressure.

```{r}
success_patients <- which(change_in_bp < 0)
success_patients
```

Alternatively we can obtain a vector of patients which saw a decrease in blood pressures as follows:

```{r}
#creates a vector subset of change_in_bp vector consisting of patients that saw a positive change, i.e. decrease in bp
positive_result <- change_in_bp[change_in_bp < 0]
positive_result
```

Lastly, we calculate the average decrease in blood pressure amongst patients who saw a decrease in blood pressure.

```{r}
avg_change_success_patients <- mean(positive_result)
avg_change_success_patients
```

So amongst those who saw a decrease in blood pressure after treatment, on average there was a decrease in -10.6 points in blood pressure. Very good!

## Task 2: Basic Data Frame Practice

First, we create a data frame consisting of data from the previous task.

```{r}
patient <- 1:20
pre_bp <- pre_treat
post_bp <- post_treat
diff_bp <- change_in_bp
bp_df <- data.frame(patient, pre_bp, post_bp, diff_bp)
bp_df
```

Next we return only the rows where the difference in blood pressures is negative. Note that this creates a new, subset data frame.

```{r}
pos_change_bp_df <- subset(bp_df, diff_bp < 0)
pos_change_bp_df
```

Now we create a new column for the original data frame which consists of TRUE if the corresponding post_bp observation less than 120, and FALSE if the corresponding post_bp observation is greater than 120.

```{r}
#creates vector of boolean values based on if post_bp elements are < 120 or not
low_bp <- bp_df$post_bp < 120

#creates new column for bp_df data frame by referencing low_bp vector as a column
bp_df$low_bp <- low_bp
bp_df
```

Lastly we print the data frame so that it looks like a nice table of data entries. 

```{r}
knitr::kable(bp_df, align = "ccccc")
```

## Task 3: List Practice

First we create a data frame similar to the one in the previous task with new data from the experiment where the treatment was just a placebo. 

```{r}
#data frame with patient number, pre "treatment" blood pressure, post "treatment" blood pressure, and the difference in blood pressure
patient <- 1:10
pre_bp <- c(138,135,147,117,152,134,114,121,131,130)
placebo_bp <- c(105,136,123,130,134,143,135,139,120,124)
diff_bp <- placebo_bp-pre_bp
placebo_df <- data.frame(patient, pre_bp, placebo_bp, diff_bp)

#new columns with TRUE/FALSE values based on post placebo blood pressure
low_bp <- placebo_df$placebo_bp < 120
placebo_df$low_bp <- low_bp
placebo_df
```

Next we create a list with two elements: the data fram from task 2 and the data frame from the previous step.

```{r}
bp_data_list <- list(treatment = bp_df, placebo = placebo_df)
bp_data_list
```

Now we will access the first list element in 3 different ways, with the first way returning a list of length 1 consisting of the data frame, and the second and third ways returning just the data frame itself. 

```{r}
bp_data_list[1]
```
```{r}
bp_data_list[[1]]
```

```{r}
bp_data_list$treatment
```

Lastly, we access the pre_bp column from the placebo data frame in the list. 

```{r}
bp_data_list$placebo[2]
```

